% =============================================================================
\section{Introduction}\label{sec:introduction}
% =============================================================================

The COVID-19 pandemic exposed the vulnerability of national health systems
to rapid, spatially uneven surges in critical-care demand. In England,
sustained demand for adult critical-care beds and mechanical ventilation
during successive waves required emergency capacity expansion and
inter-region patient transfers. Decisions of this scale are made under
deep uncertainty: planners must balance the cost of expanding surge
capacity against the risk of unmet demand, while keeping patient-transfer
burden manageable across the seven NHS England regions.
The computational intelligence community has produced strong tools for
both halves of this problem --- epidemic and hospital-demand
forecasting~\cite{gao2021stan,gao2023hoist,qian2025physics} and robust
facility-location and resource allocation under
uncertainty~\cite{eriskin2022robust,wan2023two,shamseddin2025optimization}
--- but few works connect the two into a single decision-focused pipeline
tailored to a national health system. The present work occupies the
\emph{temporal, per-region} corner of the forecasting design space and
the \emph{multi-method comparative} corner of the allocation design
space; the spatiotemporal-graph alternatives surveyed in
Section~\ref{sec:related_work} are explicitly deferred to future work.

Existing forecasting models, including physics-informed neural networks
(PINNs)~\cite{ajao2025hybrid,qian2025physics} and graph-based
architectures such as STAN~\cite{gao2021stan} and HOIST~\cite{gao2023hoist},
are typically optimised for statistical accuracy alone. Yet for
critical-care planning the operational cost of forecast errors is
asymmetric: underestimating demand by ten beds risks patients without
ventilators, while overestimating wastes prepared capacity. Standard
squared-error or absolute-error objectives treat these symmetrically.
Conversely, most optimisation formulations for pandemic capacity planning
assume demand is deterministic or drawn from off-the-shelf statistical
forecasts~\cite{shamseddin2025optimization,eriskin2022robust}, leaving
unexploited the dynamics-aware information that physics-informed
forecasters can provide. Recent work begins to close this gap:
Liu and Cao~\cite{liu2026dynamic} couple spatiotemporal graph attention
with multi-objective optimisation for vaccine and
non-pharmaceutical-intervention allocation, and Shams Eddin and El
Hajj~\cite{shamseddin2025optimization} integrate SEIRD-driven demand into
robust hospital-bed scheduling for a single US region. To our knowledge no
published work combines (a) physics-informed regional epidemic forecasting,
(b) decision-aware training that explicitly penalises operationally costly
forecast errors, and (c) metaheuristic robust optimisation for NHS England
critical-care surge capacity.

We close that gap with a forecast-to-decision pipeline organised in four
phases. \emph{Phase A} is a per-region PINN-SEIRD module that produces
multi-horizon quantile forecasts of mechanical-ventilation bed occupancy.
The PINN is first pre-trained to satisfy the refined $SEI_aI_sHCRD$ ODE
system, producing twelve time-varying physics-informed features (eight
compartmental states and four transmission/clinical parameters) that are
then fed --- frozen --- into a temporal GRU head. The head emits the
$10$th, $50$th and $90$th predictive quantiles per horizon \emph{directly}
via a multi-quantile \emph{pinball loss} (decision-aware by construction,
since the $q^{(0.9)}$ component asymmetrically penalises under-prediction
nine-to-one), with no post-hoc Monte-Carlo sampling. A learnable
level-and-trend anchor skip-connects the most recent observation and
recent weekly slope to every horizon-quantile prediction, trading
short-horizon flexibility for long-horizon stability. \emph{Phase B}
discretises the quantile forecasts into three demand scenarios
(low/median/high) with probabilities $\pi = (0.20, 0.60, 0.20)$.
\emph{Phase C} solves a regional bed-surge allocation problem using
those scenarios, comparing a deterministic mixed-integer linear
programme, a tail-risk robust MILP, a Genetic Algorithm with
LP-slave decomposition, the Non-dominated Sorting Genetic Algorithm
II~\cite{deb2002fast}, Simulated Annealing, and four closed-form
heuristic baselines. \emph{Phase D} reports operational metrics
(expected unmet, worst-case unmet, transfer burden, runtime) plus
sensitivity to the surge budget, the CVaR weight, and the mutual-aid
travel-time cap.

Our contributions are: (1) a forecast-to-decision pipeline reproducible
on open NHS data, with all four phases released as installable Python
console commands; (2) a \emph{decision-aware} multi-quantile pinball
training objective coupled with a level-and-trend anchor skip connection,
jointly responsible for a $19\%$ RMSE reduction over per-origin
re-fitted ARIMA at the operational $14$-day horizon and a four-fold
reduction in underestimation rate at $h = 28$; (3) a hybrid-regional
robust allocation model with explicit MILP/GA/NSGA-II/SA comparison,
revealing that the robust MILP achieves the lowest transfer burden in
the policy set ($4.6\times$ below population-proportional, at identical
expected and worst-case unmet) and that the metaheuristics match it
within a few percent at $20$--$115$~s runtime; (4) a refined
$SEI_aI_sHCRD$ compartmental model that separates the hospitalisation
ratio $\omega$ ($I_s\to H$) from the critical-care escalation proportion
$\phi$ ($H\to C$), correcting an over-loading of $\omega$ in earlier
work~\cite{ajao2025hybrid}; and (5) an NHS England case study spanning
five forecasters --- PinnGRU, ARIMA, per-region GRU, XGBoost, and
seasonal-naive(7) --- plus a three-axis sensitivity sweep over surge
budget, CVaR weight, and travel-time cap, quantifying when forecast
quality versus parameter choice dominates the operational outcome.
\Cref{fig:pipeline} summarises the four-phase pipeline.
\Cref{sec:related_work} reviews related work;
\Cref{sec:forecasting,sec:optimisation} describe the forecasting and
optimisation modules; \Cref{sec:case_study} presents the case study;
\Cref{sec:results,sec:discussion,sec:conclusion} report results, discuss
limitations, and conclude.

% --- Figure 1 : pipeline overview --------------------------------------------
\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  >={Stealth[length=1.6mm,width=1.2mm]},
  phase/.style={
    draw, rounded corners=2pt, line width=0.4pt,
    align=left, anchor=north west,
    inner xsep=4pt, inner ysep=4pt,
    minimum height=27mm, text width=27mm,
  },
  hdr/.style={
    font=\scriptsize\bfseries, anchor=base west,
  },
  arr/.style={->, line width=0.5pt, shorten >=1pt, shorten <=1pt},
  xsep/.style={node distance=3mm},
]
% --- four phase boxes, evenly tiled ---
\node[phase, fill=NavyBlue!6]   (A) at (0,0)
  {\textbf{A.~Forecasting}\\[1pt]
   \scriptsize NHS regional series\\
   {\scriptsize $\downarrow$ pre-train}\\
   \scriptsize PINN--SEIRD\\
   \scriptsize 8 state + 4 param\\
   \scriptsize features (frozen)\\
   {\scriptsize $\downarrow$}\\
   \scriptsize Quantile GRU head\\
   \scriptsize pinball loss\\
   \scriptsize + level/trend anchor};

\node[phase, fill=ForestGreen!6,
      right=3mm of A.north east, anchor=north west] (B)
  {\textbf{B.~Scenarios}\\[1pt]
   \scriptsize $q^{(0.1)}\!\to\!$ low\\
   \scriptsize $q^{(0.5)}\!\to\!$ median\\
   \scriptsize $q^{(0.9)}\!\to\!$ high\\[1pt]
   \scriptsize weights\\
   \scriptsize $\pi{=}(0.2,0.6,0.2)$\\[1pt]
   \scriptsize $d_{r,h}^{s}$ tensor\\
   \scriptsize shape $(R,H,S)$};

\node[phase, fill=BrickRed!6,
      right=3mm of B.north east, anchor=north west] (C)
  {\textbf{C.~Allocation}\\[1pt]
   \scriptsize Deterministic MILP\\
   \scriptsize Tail-risk robust MILP\\
   \scriptsize GA + LP slave\\
   \scriptsize NSGA-II (3-obj.)\\
   \scriptsize Simulated annealing\\[1pt]
   \scriptsize Status quo / pop /\\
   \scriptsize demand / greedy\\
   \scriptsize baselines};

\node[phase, fill=Orange!8,
      right=3mm of C.north east, anchor=north west] (D)
  {\textbf{D.~Evaluation}\\[1pt]
   \scriptsize $E[u]$, $u^{\text{worst}}$\\
   \scriptsize transfer (bed\,km)\\
   \scriptsize runtime\\
   \scriptsize Pareto front\\[1pt]
   \scriptsize Sensitivity: $B,\ \tau,\ \lambda_3$\\
   \scriptsize Forecaster sweep};

% --- arrows between phases ---
\draw[arr] (A.east) -- ($(B.west)$);
\draw[arr] (B.east) -- ($(C.west)$);
\draw[arr] (C.east) -- ($(D.west)$);

% --- arrows from D back to A and C (feedback) ---
\path (A.south west) -- (D.south east) coordinate[pos=0.5] (mid);
\draw[arr, dashed, line width=0.4pt]
  (D.south) .. controls +(0,-7mm) and +(0,-7mm) ..
  node[below, font=\scriptsize, pos=0.5] {sensitivity feedback}
  (A.south);

\end{tikzpicture}
\caption{Four-phase forecast-to-decision pipeline. Phase~A pre-trains a
per-region PINN-SEIRD on the refined $SEI_aI_sHCRD$ system and feeds its
frozen state-and-parameter features into a quantile GRU head trained with
a multi-quantile pinball loss. Phase~B discretises the
$q^{(0.1)}/q^{(0.5)}/q^{(0.9)}$ forecasts into a three-scenario demand
tensor. Phase~C compares an exact MILP, a tail-risk robust MILP,
three metaheuristics, and four heuristic baselines on the same problem
instance. Phase~D reports operational and Pareto-front metrics and
sweeps the surge budget $B$, the CVaR weight $\lambda_3$, the
travel-time cap $\tau$, and the upstream forecaster.}
\label{fig:pipeline}
\end{figure}
