% =============================================================================
\section{Related Work}\label{sec:related_work}
% Target length: ~1.25 pages (~620 words)
% Three subsections: forecasting, optimisation, forecast-to-decision pipelines.
% Anchors the positioning against Bertsimas et al. (2022) and Luo & Stellato (2024)
% and adds Valente et al. (2022) as the closest ICU-forecasting precedent and
% Kargar et al. (2024) as the augmented epsilon-constraint precedent.
% =============================================================================

The closest precedent for our pipeline is Bertsimas
et al.~\cite{bertsimas2022where}, who couple an extended DELPHI
compartmental model of COVID-19 to a bilinear mixed-integer program for the
location of one hundred mass-vaccination sites across US states, solving the
resulting non-convex problem by a coordinate-descent scheme that alternates
forward simulation of the discretised dynamics with a linear-programming
approximation of the bilinear constraints. Their pipeline is the
methodological parallel to ours; the differences are domain
(vaccination-site siting versus critical-care surge), coupling
(iterative joint optimisation versus our decoupled forecast-then-optimise),
and forecasting objective (mean-squared error on DELPHI states versus our
decision-aware multi-quantile pinball loss). Our framework also draws on
the formal Smart Predict-then-Optimize framework of Elmachtoub and
Grigas~\cite{elmachtoub2022smart} and on the neural-ODE-based equitable
opioid facility-location work of Luo and Stellato~\cite{luo2024equitable},
who fit per-state SIR-like models via a neural-ODE-inspired procedure and
embed the discretised dynamics into a McCormick-relaxed MIP. We adopt
multi-quantile training in the spirit of the COVID-19 Forecast Hub's
Weighted Interval Score evaluation~\cite{bracher2021evaluating}, which
treats the upper quantile (q$^{0.9}$) as the operationally relevant safety
bound. We review the three relevant subliteratures below.

\subsection{Epidemic and Hospital-Demand Forecasting}

Statistical and recurrent baselines---ARIMA, Prophet, and LSTM/GRU
architectures---remain competitive when applied per region~\cite{ajao2024deep}.
Closest in target to our work, Valente et al.~\cite{valente2022forecasting}
forecast COVID-19 intensive-care occupancy in the Lazio region of Italy using
a bidirectional LSTM with Google Community Mobility features lagged by
twenty-one days following Cartení et al.~\cite{carteni2020mobility}, and
exploit stochastic-ensemble Monte-Carlo dropout for prediction intervals;
they report mean absolute error of approximately seven beds on
weekly horizons. Graph neural networks add explicit spatial coupling:
STAN~\cite{gao2021stan} combines graph-attention pooling with an
SIR-consistency loss for COVID-19 case prediction at US county and state
level, and HOIST~\cite{gao2023hoist} introduces an Ising-dynamics-inspired
regulariser over an adaptively-learned location graph for hospitalisation
prediction across 2{,}299 US counties. Physics-informed neural networks
embed compartmental ordinary differential equations directly into the
training loss~\cite{ajao2025hybrid,qian2025physics}, supporting
dynamics-consistent forecasts under regime shifts. NHS-specific work is
sparse: Hunter et al.~\cite{hunter2026machine} target elective-surgery ICU
demand at a single trust, and prior work by the present
authors~\cite{ajao2024deep,ajao2025hybrid} forecasts national-scale
mechanical-ventilation demand. None of these targets the multi-region
NHS England critical-care planning problem, and all optimise statistical
rather than operationally-aligned error metrics.

\subsection{Robust Optimisation for Healthcare Resource Allocation}

Pandemic capacity planning has produced a rich optimisation literature.
Eriskin et al.~\cite{eriskin2022robust} develop a robust multi-objective
MILP for emergency-facility location and patient assignment under demand
uncertainty. Wan et al.~\cite{wan2023two} propose a two-stage
multi-objective ICU bed allocation model solved by an improved NSGA-II on
Lyon hospital data. Shams Eddin and El
Hajj~\cite{shamseddin2025optimization} integrate SEIRD-driven demand with
standard beds, buffer beds, field hospitals, and inter-hospital transfers,
validated on Northern Virginia data. In the vaccine supply-chain space,
Kargar et al.~\cite{kargar2024data} couple a Covasim agent-based
simulator~\cite{kerr2021covasim} to a multi-echelon bi-objective MILP solved
by the augmented $\varepsilon$-constraint method, exposing the
cost-versus-emissions Pareto front; we adopt the augmented
$\varepsilon$-constraint paradigm in our allocation comparison,
complementing exact MILP and NSGA-II with an exact non-dominated-solution
generator. Tang et al.~\cite{tang2025mobile} apply adaptive large
neighbourhood search to mobile-vaccination scheduling, exemplifying the
broader operational-research-meets-public-health literature beyond static
facility location. Metaheuristic methods---GA, NSGA-II~\cite{deb2002fast},
simulated annealing, and ALNS---are routinely employed when problem scale
defeats exact MILP. The bulk of this literature is single-region or
US-focused; trust-level allocation across NHS England regions and
Integrated Care Boards has, to our knowledge, not been studied with the
methodological breadth proposed here.

\subsection{Forecast-to-Decision Pipelines}

The integration of predictive forecasts into prescriptive optimisation is
a maturing research area. Elmachtoub and Grigas~\cite{elmachtoub2022smart}
formalise the Smart Predict-then-Optimize framework with a loss aware of
the downstream decision problem. In healthcare,
Bertsimas et al.~\cite{bertsimas2022where} provide the closest end-to-end
precedent via coordinate-descent coupling of DELPHI simulation with
bilinear MIP allocation; Luo and Stellato~\cite{luo2024equitable} provide
an analogous predict-then-optimise architecture for opioid
treatment-facility location with neural-ODE parameter estimation and
McCormick-envelope relaxation; and Liu and Cao~\cite{liu2026dynamic} couple
spatiotemporal graph-attention forecasting to multi-objective vaccine and
non-pharmaceutical-intervention allocation. Our framework adopts the
predict-then-optimise structure, contributes a domain-specific
decision-aware training objective that asymmetrically penalises
operationally costly under-prediction, and broadens the optimisation
comparison to include exact MILP, augmented $\varepsilon$-constraint,
NSGA-II, GA, and simulated annealing. By treating the forecasting module
as a substitutable component, the framework is forecasting-backend
agnostic: alternative graph-based or transformer-based forecasters can be
swapped in without changes to the allocation layer.
