% =============================================================================
\section{Forecasting Module}\label{sec:forecasting}
% Target length: ~2.0 pages (~900-1000 words)
% =============================================================================

The forecasting module produces multi-horizon point and quantile estimates of
critical-care demand for each NHS region. It comprises three parts: a
per-region physics-informed neural epidemic estimator, a temporal forecasting
head with a level-and-trend anchor that emits quantiles directly, and a
pinball-loss training objective that operationalises decision-aware
prediction without separate asymmetric penalty terms.

\subsection{Refined SEIRD Compartmental Model}\label{ssec:seird}

For each region $r \in \R$ with population $N_r$, we model the local epidemic
using a refined eight-compartment $SEI_aI_sHCRD$ system. Let
$U^r(t) = (S_r, E_r, I_{a,r}, I_{s,r}, H_r, C_r, R_r, D_r)$ denote
susceptible, exposed, asymptomatic infectious, symptomatic infectious,
hospitalised, critical-care, recovered, and deceased compartments
respectively. \Cref{fig:seird} shows the flow topology, in which the
distinct probabilities $\omega$ ($I_s\!\to\! H$) and $\phi$ ($H\!\to\! C$)
disambiguate the two clinical thresholds. The system evolves according to
\begin{equation}\label{eq:seird}
\begin{aligned}
\dot{S}_r     &= -\beta_r S_r (I_{s,r}+I_{a,r})/N_r + \eta_r R_r \\
\dot{E}_r     &= \beta_r S_r (I_{s,r}+I_{a,r})/N_r - \alpha E_r \\
\dot{I}_{s,r} &= \alpha\rho E_r - d_s I_{s,r} \\
\dot{I}_{a,r} &= \alpha(1-\rho)E_r - d_a I_{a,r} \\
\dot{H}_r     &= d_s\omega I_{s,r} - d_H H_r - \mu H_r \\
\dot{C}_r     &= \phi\, d_H H_r - \gamma_{c,r} C_r - \delta_{c,r} C_r \\
\dot{R}_r     &= d_s(1-\omega)I_{s,r} + d_a I_{a,r}
                + (1-\phi)d_H H_r + \gamma_{c,r}C_r - \eta_r R_r \\
\dot{D}_r     &= \mu H_r + \delta_{c,r} C_r,
\end{aligned}
\end{equation}
subject to conservation $\sum_{\bullet}\bullet_r(t)=N_r$. Here $\alpha, d_s,
d_a, d_H$ are \emph{rates} with units of inverse-days (their reciprocals are
the corresponding clinical periods), $\rho$ is the symptomatic proportion,
$\omega$ is the hospitalisation ratio of symptomatic cases, $\mu$ is the
in-hospital mortality rate, and $\beta_r, \gamma_{c,r}, \delta_{c,r}, \eta_r$
are time-varying region-specific parameters learned from data.

\paragraph{Refinement of the prior compartmental model.}
The earlier formulation in~\cite{ajao2025hybrid} overloaded the parameter
$\omega$ to represent both the $I_s \to H$ hospitalisation ratio and, via
$(1-\omega)$, the $H \to C$ critical-care escalation. These quantities are
biologically distinct: the first is determined by clinical severity at
admission, the second by in-hospital deterioration. We separate them by
introducing $\phi$ for the $H \to C$ escalation, leaving $\omega$ for
hospitalisation alone. \Cref{eq:seird} reflects this refinement.

% --- Figure 2 : SEIRD compartment graph --------------------------------------
\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  >={Stealth[length=1.6mm,width=1.2mm]},
  comp/.style={
    draw, circle, minimum size=8mm, inner sep=0pt,
    line width=0.4pt, font=\small\bfseries,
  },
  observed/.style={fill=NavyBlue!10},
  terminal/.style={fill=black!8},
  edge/.style={->, line width=0.5pt, shorten >=1pt, shorten <=1pt},
  lab/.style={font=\scriptsize, inner sep=1pt, fill=white, fill opacity=0.85, text opacity=1},
  node distance=12mm,
]
% --- top spine: S - E - I_s - H - C - (death) ---
\node[comp]                       (S)   at (0,0)         {$S$};
\node[comp, right=of S]           (E)                    {$E$};
\node[comp, right=of E]           (Is)                   {$I_s$};
\node[comp, observed, right=of Is](H)                    {$H$};
\node[comp, observed, right=of H] (C)                    {$C$};
\node[comp, terminal, right=of C] (D)                    {$D$};
% --- bottom row: I_a (split from E) and R (sink) ---
\node[comp, below=10mm of Is]    (Ia)                   {$I_{a}$};
\node[comp, terminal, below=10mm of C] (R)              {$R$};

% --- forward flow edges ---
\draw[edge] (S)  -- node[lab,above] {$\beta_r S(I_s{+}I_a)/N$} (E);
\draw[edge] (E)  -- node[lab,above] {$\alpha\rho$}              (Is);
\draw[edge] (E)  -- node[lab,right,pos=0.45] {$\alpha(1{-}\rho)$} (Ia);
\draw[edge] (Is) -- node[lab,above] {$d_s\,\omega$}              (H);
\draw[edge] (H)  -- node[lab,above] {$\phi\,d_H$}                (C);

% --- mortality from H and C ---
\draw[edge] (H.south east) .. controls +(2mm,-4mm) and +(0,8mm) ..
  node[lab,pos=0.55] {$\mu$} (D.north west);
\draw[edge] (C) -- node[lab,above] {$\delta_{c,r}$} (D);

% --- recovery into R ---
\draw[edge] (Ia) -- node[lab,below] {$d_a$} (R);
\draw[edge] (Is.south) .. controls +(0,-8mm) and +(-4mm,4mm) ..
  node[lab,pos=0.7] {$d_s(1{-}\omega)$} (R.north west);
\draw[edge] (H.south) .. controls +(-2mm,-6mm) and +(0,8mm) ..
  node[lab,pos=0.65] {$(1{-}\phi)d_H$} (R.north);
\draw[edge] (C.south) -- node[lab,right] {$\gamma_{c,r}$} (R);

% --- waning immunity R -> S ---
\draw[edge] (R.south west) .. controls +(-15mm,-6mm) and +(-2mm,-10mm) ..
  node[lab,below,pos=0.5] {$\eta_r$} (S.south);

% --- legend caption-style ---
\node[anchor=south west, font=\scriptsize, align=left]
  at ([yshift=-9mm,xshift=-5mm]Ia.south west)
  {\textbf{Observed:} $H$, $C$ (NHS daily series)
   \quad\textbf{Time-varying:} $\beta_r, \gamma_{c,r}, \delta_{c,r}, \eta_r$};

\end{tikzpicture}
\caption{Refined $SEI_aI_sHCRD$ compartmental model. The flow $E\!\to\!I_s
\!\to\! H \!\to\! C$ carries the hospitalisation ratio $\omega$ and the
critical-care escalation $\phi$ as \emph{separate} parameters --- the
former determined by severity at admission, the latter by in-hospital
deterioration. Shaded nodes ($H$, $C$) are observed compartments
constraining the PINN loss; the terminal $R$ and $D$ nodes absorb
non-observed flows. Region-specific time-varying parameters
$(\beta_r, \gamma_{c,r}, \delta_{c,r}, \eta_r)$ are emitted by the
PINN parameter network; the remaining rates are fixed clinical
constants.}
\label{fig:seird}
\end{figure}

\subsection{Per-Region PINN-SEIRD}\label{ssec:pinn}

For each region $r$, the PINN consists of two fully-connected feedforward
networks following~\cite{ajao2025hybrid}: a state network $U^r_{NN}: t
\mapsto U^r(t)$ with five hidden layers of $20$ units, and a parameter
network $X^r_{NN}: t \mapsto (\beta_r, \gamma_{c,r}, \delta_{c,r}, \eta_r)$
with three hidden layers of $20$ units. Both use tanh hidden activations,
sigmoid outputs (so all eight compartments and four parameters lie in
$[0,1]$), and Xavier initialisation. The PINN loss combines a data-fit term
on observable compartments $\{H, C\}$ --- the only ones for which NHS
England publishes daily counts at regional resolution --- with a residual
loss against \Cref{eq:seird}:
\begin{equation}\label{eq:pinn_loss}
\mathcal{L}^{PINN}_r = \!\!\sum_{t \in \T_{\text{train}}}
  \sum_{k \in \{H, C\}} \big(U^r_{NN,k}(t) - y_{r,t,k}/N_r\big)^2
+ \lambda_{ode} \!\!\sum_{t \in \T_{\text{coll}}}
  \Big\| \tfrac{dU^r_{NN}(t)}{dt} -
  \mathcal{N}\big(U^r_{NN}(t); X^r_{NN}(t)\big)\Big\|^2,
\end{equation}
where $\mathcal{N}$ denotes the right-hand side of \Cref{eq:seird}, target
observations are normalised by population, and derivatives are evaluated via
PyTorch automatic differentiation at $256$ collocation points sampled
uniformly in $\T_{\text{coll}}$ at every iteration. We pre-train each
region's PINN for $1500$ Adam steps with $\lambda_{ode}=0.1$ and learning
rate $5\!\times\!10^{-3}$; final losses converge to
$\mathcal{O}(10^{-6})$ on both terms, indicating tight fit to the
observations \emph{and} consistency with the ODE dynamics. The learned
time-varying parameters are region-specific and non-constant (supplementary
material); their predictive value is quantified by the
\emph{w/o PINN parameter features} ablation in \Cref{tab:forecast_accuracy}.

\subsection{Temporal Forecasting Head with Level-and-Trend Anchor}
\label{ssec:temporal}

Once pre-trained, the PINN is \emph{frozen}: its outputs are reused as a
physics-informed temporal embedding for the forecaster. For each lookback
window of length $L=28$ days ending at the forecast origin $t$, the
augmented feature sequence is
\begin{equation}
\mathbf{f}_{r,\tau} = \big[\,U^r_{NN}(\tau) \,\big\|\, X^r_{NN}(\tau)
                          \,\big\|\, \tilde y_{r,\tau}\,\big] \in \mathbb{R}^{13},
\quad \tau \in \{t-L+1,\dots,t\},
\end{equation}
where $\tilde y_{r,\tau} = (y_{r,\tau} - \bar y_r) / s_{y,r}$ is the
z-scored observed target on the train+val period of region $r$. The
sequence is processed by a two-layer GRU with hidden dimension $128$ and
dropout $0.15$; the last hidden state $\mathbf{h}_{r,t}$ is decoded by a
per-horizon two-layer MLP (width $64$, GELU) into a triple of standardised
quantile predictions
$\Delta_{r,t}^{h,q} \in \mathbb{R}$ for $h \in \Hset = \{7,14,21,28\}$
and $q \in \Qset = \{0.1, 0.5, 0.9\}$.

To make the GRU's job a residual one --- predicting the \emph{deviation}
from a sensible baseline rather than the absolute level --- we add a
\emph{level-and-trend anchor} skip connection:
\begin{equation}\label{eq:anchor}
\hat z_{r,t}^{h,q} \;=\; \Delta_{r,t}^{h,q}
  \;+\; \alpha^{h,q}\, \tilde y_{r,t}
  \;+\; \gamma^{h,q}\, s_{r,t}^{(7)}\,
        \!\left(\tfrac{h}{7}\right)^{\!\!\varphi^q},
\end{equation}
where $\tilde y_{r,t}$ is the most recent z-scored observation,
$s_{r,t}^{(7)} = \tilde y_{r,t} - \tilde y_{r,t-7}$ is the recent
one-week slope, and $\alpha^{h,q}, \gamma^{h,q}, \varphi^{q} \in (0,1)$
are learnable scalars (parameterised by sigmoids on free logits). The
exponent $\varphi^q$ \emph{dampens} the linear trend extrapolation per
quantile, a Holt-Winters-style mechanism that prevents the slope-based
projection from over-shooting at long horizons. We initialise
$\varphi^{q=0.5}=\sigma(2)\approx0.88$ (near-linear, accurate point
forecasting) and $\varphi^{q=0.1}=\varphi^{q=0.9}=\sigma(0)=0.5$
(square-root damping, tighter probabilistic bounds). All anchor weights
are learnt jointly with the rest of the network during forecaster training.
Point and interval forecasts in the original scale follow by inverting
standardisation: $\hat y_{r,t+h-1}^{q} = \hat z_{r,t}^{h,q}\, s_{y,r}
+ \bar y_r$.

\subsection{Decision-Aware Training via Pinball Loss}\label{ssec:loss}

The forecaster is trained by minimising the multi-quantile pinball loss
of Koenker and Bassett~\cite{koenker1978regression}, in the form
adopted by the COVID-19 Forecast Hub Weighted Interval Score
framework~\cite{bracher2021evaluating}:
\begin{equation}\label{eq:pinball}
\mathcal{L}_{\text{pinball}} \;=\; \frac{1}{|\Qset|}\sum_{q\in\Qset}
  \mathbb{E}_{(t,h)}\!\!\left[\;\max\!\big(
    q (\tilde y_{r,t+h-1}\!-\hat z_{r,t}^{h,q}),\,
    (q-1)(\tilde y_{r,t+h-1}\!-\hat z_{r,t}^{h,q})\big)\right],
\end{equation}
evaluated at all $(h,q)\in\Hset\!\times\!\Qset$ pairs. Two operational
properties are notable. First, the $q=0.5$ pinball collapses to the
absolute-error (L1) loss for the median, so it directly targets point
accuracy without an added Huber term. Second, the $q=0.9$ pinball is
\emph{asymmetric}: a forecast that falls below the realised value (an
under-prediction, $y - \hat y > 0$) incurs loss $0.9\,(y-\hat y)$, while a
forecast that exceeds it incurs only $0.1\,(\hat y - y)$ --- a nine-to-one
penalty ratio favouring conservative upper-quantile estimates. This is the
\emph{decision-aware} component, taking the role that an explicit
asymmetric hinge played in earlier compositions of the
loss~\cite{elmachtoub2022smart} but with the asymmetry built into a single
strictly proper scoring rule rather than an ad-hoc additive
term~\cite{gneiting2007strictly}. Reporting $\hat y^{0.5}$ as the point
forecast and $[\hat y^{0.1}, \hat y^{0.9}]$ as the $80\%$ predictive
interval, the same model serves point and probabilistic users with no
post-hoc Monte-Carlo sampling.

We train each region's forecaster for up to $800$ AdamW epochs
($\text{lr}=10^{-3}$, weight decay $10^{-4}$, cosine annealing), with the
training set restricted to lookback windows whose forecast targets fall
within the Alpha-wave training split. Early stopping monitors validation
pinball loss over windows whose targets fall in the Delta-wave validation
period (patience $80$). Once selected, the model is frozen and applied to
the Omicron test period via rolling-origin evaluation at $\Delta t = 7$
days, yielding $32$ origin dates over the $9$-month evaluation horizon.
