% =============================================================================
\section{Regional Allocation Model}\label{sec:optimisation}
% Target length: ~2.5-3 pages
% =============================================================================

The allocation module decides how many additional mechanical-ventilation
beds to provide in each NHS region, and how excess demand should be
redistributed to neighbouring regions through inter-regional patient
transfers, given the quantile forecasts $\hat q^{(0.1)}, \hat q^{(0.5)},
\hat q^{(0.9)}$ produced in \Cref{sec:forecasting}. We present a
deterministic mixed-integer-linear programme, a tail-risk robust
objective, and three metaheuristic comparators.

\subsection{Sets, Parameters and Decision Variables}\label{ssec:notation}

The scope is \emph{hybrid regional}: candidate sites coincide with the
seven NHS England commissioning regions, so the demand-node and
capacity-site indices collapse to a single set
$\R = \{r_1, \ldots, r_7\}$ with $|\R| = 7$. Every region is
always ``open'' (no facility-opening binary), and per-bed surge cost is
uniform across regions, so the model decides only how many surge beds to
place in each region and how to route patients between them.

Indices are: regions $r, r' \in \R$; forecast horizons
$h \in \Hset = \{7, 14, 21, 28\}$~days; demand scenarios
$s \in \Sset = \{\text{low}, \text{median}, \text{high}\}$ corresponding
to the $q^{(0.1)}, q^{(0.5)}, q^{(0.9)}$ quantile forecasts with weights
$\pi_s = (0.20, 0.60, 0.20)$. Parameters are: forecast demand
$d_{r,h}^s = \hat q_r^{(s)}(t_0 + h)$ at decision origin $t_0$; baseline
capacity $C_r$ (regional Delta-peak occupancy $\times 1.05$ safety
margin); per-region surge cap $K_r = 0.5\, C_r$; total surge budget
$B = 0.20 \sum_r C_r$; great-circle inter-centroid distance
$D_{r r'}$ in km (scaled by a 1.3 detour factor on the road network);
inter-regional travel time $T_{r r'} = 60 D_{r r'} / 80\text{ km/h}$ in
minutes; and the mutual-aid travel-time cap $\tau = 240$~min (four
hours). Decision variables are: surge-bed allocation
$b_{r,h} \in [0, K_r]$; inter-regional transfer flow
$z_{r r', h}^s \geq 0$, defined only for $(r, r')$ pairs with
$T_{r r'} \leq \tau$; and unmet demand $u_{r,h}^s \geq 0$. Auxiliary
$W \geq 0$ tracks the upper-quantile tail in the robust extension.

\subsection{Forecast-Driven Regional MILP}\label{ssec:milp}

The deterministic model places surge capacity and routes patients to
minimise unmet demand and transfer burden under the median forecast
scenario. Let $m$ denote the median scenario, corresponding to
$q^{(0.5)}$:
\begin{equation}\label{eq:deterministic_objective}
\min\;
\underbrace{\textstyle\sum_{r, h} u_{r,h}^{m}}_{\text{unmet demand}}
\;+\;\rho
\underbrace{\textstyle\sum_{r \neq r',\, h} D_{r r'}\, z_{r r', h}^{m}}_{\text{transfer burden}},
\end{equation}
where $\rho = 10^{-3}$ is a small weight that breaks ties towards
allocations using fewer long-distance transfers without distorting the
primary unmet-demand objective. The model is subject to demand
satisfaction, capacity, expansion-cap, budget, and travel-time
constraints:
\begin{align}
\textstyle\sum_{r' : T_{r r'} \leq \tau}\! z_{r r', h}^{m} + u_{r,h}^{m}
  &\geq d_{r,h}^{m}
  && \forall r \in \R,\, h \in \Hset,
  \label{eq:demand}\\
\textstyle\sum_{r' : T_{r' r} \leq \tau}\! z_{r' r, h}^{m}
  &\leq C_r + b_{r,h}
  && \forall r \in \R,\, h \in \Hset,
  \label{eq:capacity}\\
b_{r,h} &\leq K_r
  && \forall r \in \R,\, h \in \Hset,
  \label{eq:expansion}\\
\textstyle\sum_r \max_h\, b_{r,h} &\leq B,
  && \label{eq:budget}\\
z_{r r', h}^{m} &= 0
  && \text{if } T_{r r'} > \tau,
  \label{eq:travel}\\
b_{r,h},\; z_{r r', h}^{m},\; u_{r,h}^{m} &\geq 0.
\end{align}
\Cref{eq:demand} ensures that the demand $d_{r,h}^{m}$ in region $r$ at
horizon $h$ under the median scenario is either treated locally, received
from elsewhere, or recorded as unmet. \Cref{eq:capacity} caps each region's
inbound flow at its augmented capacity (baseline plus surge).
\Cref{eq:expansion} enforces the per-region expansion cap.
\Cref{eq:budget} bounds the total surge investment in the peak-horizon
demand profile (we use the peak because the same physical beds must
serve every horizon up to and including peak load).
\Cref{eq:travel} suppresses transfers exceeding the four-hour mutual-aid
window.

\subsection{Tail-Risk Robust Objective}\label{ssec:robust}

To explicitly hedge against the upper-quantile demand realisation
($s = \text{high}$, $q^{(0.9)}$), the robust model generalises the
median formulation by imposing the demand and capacity constraints for
every $s \in \Sset$ and adding a tail-risk penalty. Define the tail
scenario subset
$\Sset_{\mathrm{tail}} = \{\text{high}\}$ and the auxiliary
$W \geq \sum_{r, h} u_{r,h}^s$ for every
$s \in \Sset_{\mathrm{tail}}$. The robust objective is
\begin{equation}\label{eq:robust_objective}
\min\;
\sum_{s \in \Sset} \pi_s
\Big(
\textstyle\sum_{r, h} u_{r,h}^s + \rho \sum_{r \neq r',\, h} D_{r r'}\,
z_{r r', h}^s
\Big)
\;+\;\lambda_3\, W,
\end{equation}
where $\lambda_3 = 0$ recovers the risk-neutral three-scenario
objective and $\lambda_3 \to \infty$ yields a worst-case (min-max)
policy on the high-demand scenario. We report results for
$\lambda_3 = 1$, giving equal weight to expected shortage and
high-scenario tail shortage.

\subsection{Exact and Metaheuristic Solvers}\label{ssec:solvers}

For the regional formulation the model has 28 surge-bed variables, up
to $7 \times 7 \times 4 \times 3 = 588$ transfer-flow variables, and
$7 \times 4 \times 3 = 84$ unmet-demand variables, plus the tail
auxiliary in the robust extension. We compare an exact MILP, three
metaheuristics, and four naive baselines on the same problem instance.

\paragraph{Exact MILP via PuLP+CBC.}
The deterministic and robust models are formulated in PuLP and solved
with the bundled CBC branch-and-cut solver. The deterministic model
optimises only the median scenario by construction, and the achieved
allocation is then re-evaluated by an LP slave over the full three-
scenario set to obtain honest expected- and worst-case unmet metrics.
The robust model optimises all three scenarios with the
$\lambda_3 = 1$ tail-risk term active. Both runs finish in well under a
second on the seven-region instance.

\paragraph{Master--slave structure for the metaheuristics.}
For fixed peak surge allocation $b^\star_r = \max_h b_{r,h}$, the
residual problem in $(b, z, u)$ becomes a linear programme on the
transfer-flow and unmet-demand variables alone. Every metaheuristic
operates on the continuous vector $\mathbf{b}^\star \in \mathbb{R}_+^7$
subject to $\sum_r b^\star_r \leq B$ and $b^\star_r \leq K_r$, calling
the LP slave once per fitness evaluation to compute the resulting
expected unmet, transfer burden, and worst-case unmet.

\paragraph{Genetic Algorithm (single-objective).}
A pymoo Genetic Algorithm minimises expected unmet (plus the tie-break
transfer-burden term) over $\mathbf{b}^\star$ with real-valued encoding,
SBX crossover ($\eta_c = 15$, $p_c = 0.9$), polynomial mutation
($\eta_m = 20$, $p_m = 1/7$), tournament selection, population size 50,
40 generations. A repair operator projects each candidate onto the
budget simplex $\sum_r b^\star_r \leq B$ after every genetic operation,
following Resende and Werneck~\cite{resende2004hybrid}.

\paragraph{NSGA-II (three-objective).}
The Non-dominated Sorting Genetic Algorithm II~\cite{deb2002fast}
produces a Pareto front over three objectives evaluated simultaneously:
\begin{align}
f_1(\mathbf{b}^\star) &= \textstyle\sum_r b^\star_r,
  &&\text{(total surge beds, ``cost'')} \label{eq:nsga2_f1}\\
f_2(\mathbf{b}^\star) &= \textstyle\sum_{r,h,s} \pi_s u_{r,h}^{s\,\star},
  &&\text{(expected unmet demand)} \label{eq:nsga2_f2}\\
f_3(\mathbf{b}^\star) &= \textstyle\sum_{r \neq r',h,s}
  \pi_s D_{r r'} z_{r r', h}^{s\,\star},
  &&\text{(transfer burden)} \label{eq:nsga2_f3}
\end{align}
where the starred quantities are LP-slave optima for the given
$\mathbf{b}^\star$. We use pymoo's default NSGA-II operators with the
same encoding, population, and generation count as the single-
objective GA. A \emph{representative point} is selected from the
non-dominated set by the L2-distance-to-ideal-point rule, normalising
each objective to $[0, 1]$ on the front.

\paragraph{Simulated Annealing comparator.}
A hand-rolled Simulated Annealing run uses geometric cooling
($T_0 = 1.0$, $\gamma = 0.97$) over Gaussian-perturbation neighbourhoods
($\sigma = 0.10\, B$) with budget-simplex repair after every proposal,
sharing the same LP-slave fitness oracle. The run length is 400
proposals.

\paragraph{Heuristic baselines.}
For completeness we include four closed-form allocation policies:
\emph{status quo} ($\mathbf{b}^\star = 0$, no surge); \emph{population-
proportional}, which distributes the budget by ONS region population;
\emph{demand-proportional}, which distributes by the expected peak
demand $\sum_{h,s} \pi_s d_{r,h}^s$; and \emph{greedy shortage-first},
which iteratively assigns single-bed increments to the region with the
currently highest LP-slave unmet demand. Each closes
$\mathbf{b}^\star$ via the same LP slave, providing a no-optimisation
performance floor.
