% =============================================================================
\section{NHS England Case Study}\label{sec:case_study}
% Target length: ~1 page (~450-500 words)
% =============================================================================

\subsection{Data and Spatial Resolution}

The case study uses two publicly available data sources covering
$1$~August~$2020$ to $31$~August~$2022$, the window during which NHS
England published daily regional admissions and bed-occupancy data:
NHS~England COVID-19~Hospital~Activity~\cite{nhs_covid_activity}, which
provides daily new admissions, total hospital cases, total occupied beds,
and mechanical-ventilation bed occupancy at NHS regional resolution; and
ONS Mid-Year Population Estimates~\cite{ons_mye_2021} on the 2021
geography, which provide the regional population denominators used both
by the per-region PINN to normalise hospital and critical-care
compartments to per-capita scale and by the population-proportional
allocation baseline. All experiments use the seven NHS England
commissioning regions: London (Y56), South West (Y58), South East (Y59),
Midlands (Y60), East of England (Y61), North West (Y62), and North East
and Yorkshire (Y63). Inter-regional travel times $T_{r r'}$ are
approximated by great-circle distances between regional centroids
scaled by a $1.3$ road-detour factor and divided by an $80$~km/h motorway
average speed.

\subsection{Train, Validation, and Test Splits}

The series is split chronologically by epidemic wave, deliberately
holding the most clinically distinct wave in the test set: training on
$1$~Aug~$2020$ -- $31$~May~$2021$ (Alpha and early vaccination),
validation on $1$~Jun -- $30$~Nov~$2021$ (Delta) for hyper-parameter
selection and early stopping, and testing on $1$~Dec~$2021$ --
$31$~Aug~$2022$ (Omicron and post-Omicron). Omicron is held out because
its case-to-hospitalisation and hospitalisation-to-critical-care ratios
differ markedly from earlier waves, providing a stress test of the
framework's generalisation under regime shift.

\subsection{Allocation Problem Parameters}\label{ssec:case_params}

Baseline regional capacity $C_r$ is set to the Delta-wave peak
mechanical-ventilation occupancy of each region, scaled by a $1.05$
safety margin, giving an operationally credible ceiling against which
the Omicron surge can be planned. The per-region surge cap is
$K_r = 0.5\, C_r$; the total surge budget is $B = 0.20 \sum_r C_r =
213$ beds; the mutual-aid travel-time cap is $\tau = 240$~min
(four hours), the operational limit reported for inter-regional
critical-care transfers during the COVID-19 response. Within the
objective, the unmet-demand term is the primary minimiser and the
transfer-burden term is a tie-break with weight $\rho = 10^{-3}$,
chosen so that one bed-day of unmet demand outweighs $1000$~bed-km of
transfer movement. The CVaR weight defaults to $\lambda_3 = 1$ for the
robust formulation, and the three demand scenarios use weights
$\pi = (0.20, 0.60, 0.20)$ on $q^{(0.1)}, q^{(0.5)}, q^{(0.9)}$
respectively. All these defaults are swept individually in the E6
sensitivity analysis of \Cref{ssec:e5_e6}.

The proposed forecaster and all baselines run with fixed random seeds;
forecast metrics are macro-averaged across the seven regions and the
$32$ rolling-origin evaluations spanning the Omicron test period.
Allocation experiments are reported at the first Omicron decision
origin ($29$~Dec~$2021$), the earliest origin with full region-by-horizon
forecast coverage. Code, configurations, and the cleaned regional
dataset are available at the repository linked in the acknowledgements.
