% =============================================================================
\section{Conclusion}\label{sec:conclusion}
% =============================================================================

We presented a forecast-to-decision pipeline for NHS critical-care
surge planning that couples a per-region physics-informed forecaster
to a tail-risk robust allocation linear program. Matching the
forecaster's loss to the operational cost of under-prediction is what
makes it useful. PinnGRU wins the short-to-medium horizons that drive
surge decisions, cutting $14$-day RMSE by $19\%$ over re-fitted ARIMA
and $37\%$ over a neural GRU, and we report rather than hide that
ARIMA is stronger at the longest horizon. On the decision side the
gain is structural. Once the budget binds, every budget-exhausting
policy ties on coverage and the robust formulation wins by routing the
same surge with the least patient transfer, while the
deterministic-to-robust gap measures what scenario-weighted
expectation buys over median-only planning. The principal caveats are
detour-scaled travel times, well-mixed regional populations, scaled
bed and bed-km cost units, and a single realised allocation origin.
The clearest next step is a trust-level instance with hundreds of
sites, where exact solution becomes expensive and a metaheuristic
comparison regains its purpose, alongside age-stratified pathways and
all-origin allocation distributions. The wider contribution is a
reproducible template for tying cost-asymmetric epidemic forecasting
to transparent and robust capacity planning on open NHS data.
